---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Forecast case and sequence notifications using variant of concern strain dynamics

[![R-CMD-check](https://github.com/epiforecasts/forecast.vocs/workflows/R-CMD-check/badge.svg)](https://github.com/epiforecasts/forecast.vocs/actions/workflows/R-CMD-check.yaml) [![Codecov test coverage](https://codecov.io/gh/epiforecasts/forecast.vocs/branch/main/graph/badge.svg)](https://app.codecov.io/gh/epiforecasts/forecast.vocs)

![Universe](https://epiforecasts.r-universe.dev/badges/forecast.vocs) [![MIT license](https://img.shields.io/badge/License-MIT-blue.svg)](https://github.com/epiforecasts/forecast.vocs/blob/master/LICENSE.md/) [![GitHub contributors](https://img.shields.io/github/contributors/epiforecasts/forecast.vocs)](https://github.com/epiforecasts/forecast.vocs/graphs/contributors) 


Contains models and tools to produce short-term forecasts for both case and sequence notifications assuming circulation of either one or two variants. Tools are also provided to allow the evaluation of the use of sequence data for short-term forecasts in both real-world settings and in user generated scenarios.

## Installation

Install the stable development version of the package with:

```{r, eval = FALSE}
install.packages("forecast.vocs", repos = "https://epiforecasts.r-universe.dev")
```

Install the unstable development from GitHub using the following, 

```{r, eval = FALSE}
remotes::install_github("epiforecasts/forecast.vocs", dependencies = TRUE)
```

Install stan to enable model fitting and forecasting using:

```{r, eval = FALSE}
cmdstanr::install_cmdstan()
```

## Quick start

This quick start uses data from Germany that includes COVID-19 notificatons and sequences with sequences either being positive or negative for the Delta variant. It shows how to produce forecasts for both a one and two strain model for the 19th of June 2021 when the latest available data estimated that approximately 7% of COVID-19 were positive for the Delta variant. Note that estimated growth rates and reproduction numbers shown here have been rescaled using an assumed generation time of 5.5 days.

```{r}
library(forecast.vocs)
options(mc.cores = 4)

obs <- filter_by_availability(
  germany_covid19_delta_obs,
  date = as.Date("2021-06-19")
)

current_obs <- latest_obs(germany_covid19_delta_obs)
```

### Forecast all-in-one

Run a forecast for both one and two strain models (or optionally just one of these) using the `forecast()` function. This provides a wrapper around other package tooling to initialise, fit, and summarise forecasts. Multiple forecasts can be performed efficiently across dates and scenarios using `forecast_across_dates()` and `forecast_accross_scenarios()`.

```{r, message = FALSE}
forecasts <- forecast(obs,
  strains = c(1, 2), voc_label = "Delta", scale_r = 5.5 / 7,
  adapt_delta = 0.99, max_treedepth = 15,
  refresh = 0, show_messages = FALSE,
)
forecasts
```

Unnest posterior estimates from each model.

```{r}
posteriors <- unnest_posterior(forecasts)
```

Plot the posterior prediction for cases from the single strain model ("Overall"), from the two strain model ("Combined"), and the unobserved estimates for each strain.

```{r}
plot_cases(posteriors, current_obs)
```

Plot the posterior prediction for the fraction of cases that have the Delta variant from the two strain model.

```{r}
plot_voc(posteriors, current_obs, voc_label = "Delta variant")
```

Plot the posterior estimate for the growth rate over the mean of the generation time for COVID-19 cases (here assumed to be 5.5 days).

```{r}
plot_growth(posteriors)
```

Plot the posterior estimate for the effective reproduction number of Delta and non-Delta cases.

```{r}
plot_rt(posteriors)
```

Alternatively a list of plots with sensible defaults can be produced using the following. 

```{r, eval = FALSE}
plot_posterior(
  posteriors, current_obs,
  voc_label = "Delta variant"
)
```

### Step by step forecast

Rather than using the all-in-one `forecast()` function inidividual package functions can be used to produce a forecast as follows.

```{r, eval = FALSE}
dt <- stan_data(obs, horizon = 4)

model <- load_model(strains = 2)

inits <- stan_inits(dt, strains = 2)

fit <- stan_fit(
  data = dt, model = model, init = inits,
  adapt_delta = 0.99, max_treedepth = 15,
  refresh = 0, show_messages = FALSE
)

posterior <- summarise_posterior(fit, voc_label = "Delta", scale_r = 5.5 / 7)
```

As for forecasts produced with the `forecast()` function summary plots and estimates can then be produced using package functions.

## Citation

If using `forecast.vocs` in your work please consider citing it using the following,

```{r, echo = FALSE}
citation("forecast.vocs")
```

## How to make a bug report or feature request

Please briefly describe your problem and what output you expect in an [issue](https://github.com/epiforecasts/forecast.vocs/issues). If you have a question, please don't open an issue. Instead, ask on our [Q and A page](https://github.com/epiforecasts/forecast.vocs/discussions/categories/q-a).

## Contributing

We welcome contributions and new contributors! We particularly appreciate help on priority problems in the [issues](https://github.com/epiforecasts/forecast.vocs/issues). Please check and add to the issues, and/or add a [pull request](https://github.com/epiforecasts/forecast.vocs/pulls).

## Code of Conduct
  
Please note that the `forecast.vocs` project is released with a [Contributor Code of Conduct](epiforecasts.io/forecast.vocs/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
