---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Evaluating the impact of modelling strain dynamics on short-term COVID-19 forecast performance

[![R-CMD-check](https://github.com/epiforecasts/bp.delta/workflows/R-CMD-check/badge.svg)](https://github.com/epiforecasts/bp.delta/actions/workflows/R-CMD-check.yaml) [![paper](https://github.com/epiforecasts/bp.delta/workflows/paper/badge.svg)](https://github.com/epiforecasts/bp.delta/actions/workflows/paper.yaml)

This repository contains stan code for one and two strain branching process process models as well as an R package for using these models to produce forecasts of COVID-19 notified cass in Germany. See the paper for further method details and the rest of this README for installation and example functionality.

## Installation

Either install the package from GitHub using the following, 

```{r, eval = FALSE}
devtools::install_github("seabbs/bp.delta", dependencies = TRUE)
```

Or install the package locally (with the working directory at the package root),

```{r, eval = FALSE}
devtools::install_dev_deps()
devtools::load_all()
```

Install stan using:

```{r, eval = FALSE}
cmdstanr::install_cmdstan()
```

## Using the model

### Step by step forecast

*Note: Stan throws some initialisation warnings at the moment and a low number of divergent transitions (<0.5%) regardless of auto-regressive model formulation or stan settings. This is a work in progress - remove the `refresh = 0, show_messages = FALSE ` settings to see these issues.*

```{r}
library(bp.delta)
options(mc.cores = 4)

obs <- latest_obs(germany_obs)

dt <- stan_data(obs, horizon = 4)

model <- load_model(strains = 2)

inits <- stan_inits(dt, strains = 2)

fit <- stan_fit(
  data = dt, model = model, init = inits,
  adapt_delta = 0.99, max_treedepth = 15,
  refresh = 0, show_messages = FALSE
)

posterior <- summarise_posterior(fit)
```

Plot the posterior prediction for cases.

```{r}
plot_cases(posterior, obs, max(obs$date), log = TRUE)
```

Plot the posterior prediction for the fraction of cases that are DELTA.

```{r}
plot_delta(posterior, obs, max(obs$date))
```

Plot the posterior estimate for the effective reproduction number of DELTA and non-DELTA cases.

```{r}
plot_rt(posterior, max(obs$date))
```

### Forecast wrapper

Run a complete forecast for both the one and two strain models using the `forecast` function (change the `save_path` argument to alter the location where results are saved). See `names(results)` for a breakdown of the output (including summarised posteriors and plots).

```{r, message = FALSE}
results <- forecast(obs,
  strains = c(1, 2),
  adapt_delta = 0.99, max_treedepth = 15,
  refresh = 0, show_messages = FALSE
)
```

Plot the posterior prediction for cases for both models.

```{r}
results$plots$log_cases
```

Plot the posterior estimate for the effective reproduction number of DELTA, non-DELTA cases, and overall.

```{r}
results$plots$rt
```

### Further details

See the following for more detailed analysis:

- `inst/analysis/germany/forecast.R:` for a forecasting application.
- `inst/analysis/germany/retrospective.R` for retrospective fitting and model evaluation
