---
title: Model definitions
author: Sam Abbott
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Model definitions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

The package currently supports two models both with various options. A baseline single strain model and a two strain model with flexible variable dynamics. These are defined in detail below.

## Single strain

We model the mean ($\lambda_t$) of reported cases ($C_t$) as an order 1 autoregressive (AR(1)) process on the log scale by time unit ($t$). The model is initialised by assuming that the initial reported cases are representative with a small amount of error (2.5%).

\begin{align}
  \log \lambda_0 &\sim \text{LogNormal}\left(\log C_0 , 0.025 \times \log C_0 \right) \\
  \log \lambda_t &= \log \lambda_{t-1} + r_t
\end{align}

Where $r_t$ can be interpreted as the growth rate and the exponential of $r_t$ as the effective reproduction number ($R_t$) assuming a mean generation time equal to the scaling rate used (see the documentation for `scale_r` in `?forecast()`). $r_t$ is then itself modelled as a differenced AR(1) process, 

\begin{align}
  r_0 &\sim \text{Normal}\left(0, 0.25 \right) \\
  r_t &= r_{t-1} + \epsilon_t  \\
  \epsilon_0 &= \eta_0 \\
  \epsilon_t &= \beta \epsilon_{t-1} + \eta_t
\end{align}

Where, 

\begin{align}
  \eta_t &\sim \text{Normal}\left(0, \sigma \right) \\
  \sigma &\sim \text{Normal}\left(0, 0.1 \right) \\
  \beta &\sim \text{Normal}\left(0, 1 \right)
\end{align}

We then assume a negative binomial observation model with overdispersion $\phi_c$ for reported cases ($C_t$),

\begin{align}
  C_{t} &\sim \text{NegBinomial}\left(\lambda_t, \phi_c\right) \\
  \frac{1}{\sqrt{\phi_c}} &\sim \text{Normal}(0, 1) 
\end{align}

Where $\sigma$, and $\frac{1}{\sqrt{phi_c}}$ are truncated to be greater than 0 and $\beta$ is truncated to be between -1 and 1. Optionally a Poisson observation model may instead be used (see the documentation for `overdispersion` in `?forecast()`).

The stan code for this model available [here](https://github.com/epiforecasts/forecast.vocs/blob/main/inst/stan/bp.stan)  or can be loaded into `R` using the following code,

```{r}
library(forecast.vocs)
readLines(fv_model(strains = 1, compile = FALSE))
```

## Two strain

We model strain dynamics using the single strain model as a starting point but with the addition of strain specific AR(1) variation and a beta binomial (or optionally binomial) observation process for sequence data. The full two strain model is described below. Parameters related to the variant of concern (VoC) are given the $\delta$ superscript and parameters related to non-VoC cases are given the $o$ superscript.

Mean reported cases are again defined using a AR(1) process on the log scale for each strain and the combined for overall mean reported cases. 

\begin{align}
  \log \lambda_0 &\sim \text{LogNormal}\left(\log C_0 , 0.025 \times \log C_0 \right) \\
  \log \lambda^{\delta}_0 &\sim \text{LogNormal}\left(\log C^{\delta}_0 , 0.025 \times \log C^{\delta}_0 \right) \\
  \log \lambda^{o}_0 &= \log \left(\lambda_0 - \lambda^{\delta}_0 \right) \\
  \log \lambda^{\delta}_t &= \log \lambda^{\delta}_{t-1} + r^{\delta}_t \\
  \log \lambda^{o}_t &= \log \lambda^{o}_{t-1} + r^{o}_t \\
  \lambda_t &= \lambda^{\delta}_t + \lambda^{o}_t
\end{align}

Where $C^{\delta}_0$ is derived by calculating the mean proportion of cases that had the VoC for the first time point using the overall number of reported cases, the number of sequenced cases, and the number of sequences that were positive for the VoC. The growth rate for VoC and non-VoC cases ($r^{o, \delta}_t$) is then modelled as a combination of an overall growth rate ($r_t$ as defined for the single strain model), a strain specific modifier ($s^{o, \delta}_0$), and an AR(1) error term post introduction ($\epsilon^{o, \delta}_t$).

\begin{align}
  r^{o, \delta}_t &= r_t + s^{o, \delta} + \epsilon^{o, \delta}_t \\
  \epsilon^{o, \delta}_0 &= 0 \\
  \epsilon^{o, \delta}_t &= \epsilon^{o, \delta}_{t-1} + \eta^{o, \delta}
\end{align}

Where, 

\begin{align}
  \eta^{o, \delta}_t &\sim \text{Normal}\left(0, \sigma^{o, \delta} \right) \\
  \sigma^{o, \delta} &\sim \text{Normal}\left(0, 0.1 \right) \\
  s^o &= 0 \\
  s^{\beta} &\sim \text{Normal}(0, 0.5)
\end{align}

Which assumes an uninformed transmissibility difference for the VoC vs non-VoC cases. It is strongly encouraged that users use their domain knowledge to inform this prior. The relationship between variations can be altered so that the VoC differs only by the transmissibility difference (s^{\beta}) to the non-VOC strain and any variation over time is shared between strains. Optionally strains can also be modelled to evolve independently over time. See the documentation for `variant_relationship` in `?forecast()` for details. For most users the default pooled method is recommended. 


Finally, the mean proportion of samples that have the VoC ($p_t$) is then estimated using the mean reported cases with the VoC and the overall mean reported cases.

\begin{equation} 
  p_t = \frac{\lambda^{\delta}_t}{\lambda_t}
\end{equation}

We assume a beta binomial observation model for the number of sequences ($N_t$) that are positive ($P_t$) for the VoC with overdispersion $\phi_s$.

\begin{align}
  P_{t} &\sim \mathrm{BetaBinomial}\left(N_t, p_t \phi_s, (1 - p_t) \phi_s\right) \\
  \frac{1}{\sqrt{\phi_s}} &\sim \mathcal{N}(0, 1) 
\end{align}

Where $\sigma^{o, \delta}$, and $\frac{1}{\sqrt{\phi_s}}$ are truncated to be greater than 0. A binomial observation model is also available (see the documentation for `overdispersion` in `?forecast()`).

The stan code for this model available [here](https://github.com/epiforecasts/forecast.vocs/blob/main/inst/stan/twostrainbp.stan) or can be loaded into `R` using the following code,

```{r}
library(forecast.vocs)
readLines(fv_model(strains = 2, compile = FALSE))
```
