---
title: "paper"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{paper}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: library.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
header-includes:
   - \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE,
                      fig.width = 9, fig.height = 9,
                      dpi = 320,
                      fig.path = "figures/")
```


```{r, message = FALSE}
library(bp.delta)
```


## Authors

S. Abbott (1), J. Bracher (2), S. Funk (1), ... (TBD)

Correspondence to: sam.abbott@lshtm.ac.uk

## Affiliations

1. Center for the Mathematical Modelling of Infectious Diseases, London School of Hygiene & Tropical Medicine, London WC1E 7HT, United Kingdom
2. Chair of Statistics and Econometrics, Karlsruhe Institute of Technology (KIT), Karlsruhe, Germany

## Abstract

**Background:** 

**Methods:** 

**Results:** 

**Conclusions:** 

## Introduction

### Why

### What

## Methods

### Data

### Models

#### Single strain

#### Two strain#

#### Fitting

### Evaluation

### Data availability scenarios

### Reproducibility

All analysis was carried out using R version 4.0.5 [@R]. The analysis is available as an R package. A dockerfile has been made available with the code to enhance reproducibility [@Boettiger:2015dw].

## Results

## Discussion

**Summary**

**Strengths and weaknesses**

**Literature comparison**

**Further work**

**Conclusions**

**Data availability**

Zenodo:

This project contains the following underlying data:


License: MIT

**Software availability**

Source code is available from: 

Archived source code at time of publication: 

License: MIT

**Contributors**

SA conceived and designed the work. SA undertook the analysis with advice from all other authors. All authors wrote the first draft of the paper and all authors contributed to subsequent drafts. All authors approve the work for publication and agree to be accountable for the work.

**Grant information**

This work was supported by the Wellcome Trust through a Wellcome Senior Research Fellowship to SF [210758].

**Competing interests**

There are no competing interests. 


## References

<div id = 'refs'></div>